Placeholder for article 3

### 2025-01-22 analysis/paper/paper.qmd
Improved results main text for class/type and retouched tool tables.

The final lithic sample from the Proto-Solutrean assemblage, consisting of the levels 4E and top of level 5, consists of 635 artifacts. When crossing the lithic class with chert type, we see that the local types show high frequency of blanks (\>50%, including blank fragments), followed by shatter (\~11-27%), cores (\~9-12%) and retouched tools (\~5-15%). A single burin spall was identified in T4. There is also the absence of core preparation and maintenance products across all local cherts. T1 (chalcedony) is composed mostly of blanks and a retouched tool (Vale Comprido point). The absence of cores but presence of shatter may hint at the knapping of this raw material at the site, but the absence of formal cores perhaps due to the quality of nodules/lenses, which are characterised by fractures. For non-local types, especially those with higher number of artefacts (T6-T8), blanks and blank fragments are the most frequent class (48.3%, 53.1% and 69.2%, respectively). Retouched tools are present in smaller frequencies for the previous mentioned chert types (\~17-23%). For T6 the absence of cores but the presence of shatter and a single core preparation and maintenance product suggests knapping on site. Similarly, T7 shows the presence of cores and core fragments (8.4%), shatter (17.0%) and a single core preparation and maintenance product. In comparison, T8, despite more limited in number (n=13), shows only one core, with the rest of the artefacts being blanks (as previously noted) or retouched tools (23.1% and n=3). Aside from the TL artefacts, all non-local cores show at least 1 core and frequently the presence of shatter. As expected, the TL artifacts include a blank and a retouched. However, a shatter was also identified. In total, 65 chert retouched tools were identified, with 23.1% (n=15) having an unidentified chert type. A total of 26 retouched tools correspond to local cherts, while 23 correspond to non-local cherts.

@@

Regarding the typology, 14 different types of retouched tools were identified. Endscrappers (26.2%), splintered pieces (17%), burins (12.3%) and retouched flakes (7.7%) were the most prevalent. Vale Comprido points, the Proto-Solutrean fossil director, are barely present in chalcedony and chert (n=1), as they are mostly present in dolerite, while chert was mainly used for other tools as seen in previous studies (Belmiro et al., 2021). Regarding the types of chalcedony and chert, local raw materials show 9 different typologies, with Vale Comprido points, Solutrean retouch and notches being present only in the local types, even if in small numbers. Non-local cherts show 10 different typologies of retouched tools. Perforator-endscrapper, truncation, backed bladelet and retouched blades typologies can only be found in non-local varieties, albeit also in small numbers.

@@

The Solutrean assemblages show similar patterns to those observed in previous assemblages. The final sample from levels A, B and C from the Solutrean occupations of the Shelter area consists of 914 artifacts. When crossing the class data with type of raw material, we see that local chert, especially T2 (n=506) which accounts for the majority of local chert present in the studied samples, the majority of the debitage is composed of flakes and elongated blanks (83.9%). Cores are also present (10.5%), alongside a small percentage of core preparation and maintenance products (1.3%). In comparison to debitage products, retouched tools are scarce in the sample (\~3%). The non-local cherts most present in the Solutrean of the Shelter area are T6 (n=86) and T8 (n=30). These show similar patterns to the local cherts, with high percentages of blanks (\>50%). Cores and core preparation and maintenance are also present in small percentages (\~3-6% and 3.5% only for T6, respectively), as well as retouched tools (\<10%). However, these specific results are truncated by the lack of shatter, which were not present in the sample used for the study (see Materials and Methods section). It is important to notice, however, that alike previous results from the Solutrean assemblages of the Terrace area, a large portion of the studied sample was altered and did not allow for a chert type attribution (21.9912473%).

@@

The analysed Solutrean sample has a total of 43 retouched tools (complete and fragments), with a high variety of retouched tools (11), ranging from 5 (11.6%) to 4 (9.3%) per tool type. These results however, are significantly reduced from the original dataset (Cascalheira et al., 2013), possibly due to no bucket coordinates being used. The typologies which are most frequently present in the studied sample are bifacial blanks and bifacial fragments, amounting to a total of 23.2% of products related to bifacial tool/point production. Two types of points were identified: shouldered points (2 complete and 1 fragment, amounting to a total of 6.9%) and tanged and 4 winged points (9.3%). Other frequent retouched typologies are end scrappers (9.3%), notches (11.6%), retouched blanks (\~14%) and splintered pieces (11.6%). Regarding chert type, a substantial portion of retouched tools are of unidentified chert types (n=13), mostly due to alterations (fire and post depositional). Most retouched tools with identified chert are local (n=13, accounting to 30.2325581%), followed by Type 6 cherts, previously suggested to be from Central Portugal (n=7, accounting to 16.2790698%). Local types show a higher variability retouched typologies (9 types), although this may be related to the higher number of tools. In comparison, Type 6 shows only 5 different typologies.

>>>>

The final lithic sample from the Proto-Solutrean assemblage, consisting of the levels 4E and top of level 5, consists of 635 artifacts. Of these, 167 (`r round(167*100/635, 2)`%) have an indeterminate chert type. Local cherts correspond to `r 32+62+32+158+51` artefacts (`r round((32+62+32+158+51)*100/635, 2`%), with T4 being the most present local chert (n=158 or `r round(158*100/(32+62+32+158+51), 2)`% of the local types). Non-local types are present in smaller numbers (n=`r 60+47+13+9+1` or `r round((60+47+13+9+1)*100/635, 2)`%), with T6 and T7 being the most relevant (n=60 and n=47, respectively), and the TL group includes 3 artefacts.

When crossing the lithic class with chert type (@tbl-general-proto), we see that the local types show high frequency of blanks (\>50%, including blank fragments), followed by shatter (\~11-27%). Cores are present in small numbers (n=`r 4+5+13+6`), corresponding to ~9-12% of the classes in each local type, with the exception of T1 (chalcedony), with no cores. Core preparation and maintenance products are absent in all local cherts. For non-local types, especially those with higher number of artefacts (T6-T8), blanks and blank fragments are the most frequent class (48.3%, 53.1% and 69.2%, respectively). Retouched tools are present in smaller frequencies for the previous mentioned chert types (\~17-23%). For T6 the absence of cores but the presence of shatter and a single core preparation and maintenance product suggests knapping on site; all other non-local types show at least 1 core and frequently the presence of shatter. As expected, the TL artifacts include a blank and a retouched. However, a shatter was also identified.

@@

In total, 65 chert retouched tools were identified, with 23.1% (n=15) having an unidentified chert type (@tbl-retouch-proto). A total of `r 1+10+2+8+5` (`r round((1+10+2+8+5)*100/65, 2)`%) retouched tools correspond to local cherts, while `r 10+9+3+1+1` (`r round((10+9+3+1+1)*100/65, 2)`%) correspond to non-local and TL cherts. As seen in @tbl-retouch-proto, 14 different types of retouched tools were identified. Endscrappers (26.2%), splintered pieces (17%), burins (12.3%) and retouched flakes (7.7%) were the most prevalent. Local raw materials show 9 different typologies, with Vale Comprido points, Solutrean retouch and notches being present only in the local types, even if in small numbers. Non-local cherts show 10 different typologies of retouched tools. Perforator-endscrapper, truncation, backed bladelet and retouched blades typologies can only be found in non-local varieties, albeit also in small numbers.

@@

The Solutrean assemblages show similar patterns to those observed in previous assemblages. The final sample from levels A, B and C from the Solutrean occupations of the Shelter area consists of 914 artifacts. Of these, 201 (`r round(201*100/914, 2)`%) correspond to the indeterminate chert group. Local cherts correspond to `r 16+506+4+10+30` (`r round((16+506+4+10+30)*100/914, 2)`%) artefacts, mostly from T2 (n=506). Non-local cherts (including the TL group) correspond to `r 86+9+30+3+1+17+1` (`r round((86+9+30+3+1+17+1)*100/914, 2)`%) artefacts, mostly from T6 (n=86).

When crossing the class data with type of raw material (@tbl-general-solutrean), we see that for local chert, especially T2, the majority of the debitage is composed of flakes and elongated blanks (83.9%). Cores are also present (10.5%), alongside a small percentage of core preparation and maintenance products (1.3%). In comparison to debitage products, retouched tools are scarce in the sample (\~3%). Non-local cherts, especially T6 and T8, show similar patterns to the local cherts, with high percentages of blanks (\>50%). Cores and core preparation and maintenance are also present in small percentages (\~3-6% and 3.5% only for T6, respectively), as well as retouched tools (\<10%). However, these specific results are truncated by the lack of shatter, which were not present in the sample used for the study (see Materials and Methods section).

@@

The analysed Solutrean sample has a total of 43 retouched tools (complete and fragments), with `r 3+13+1` corresponding to local cherts and `r 7+1+2+3` to non-local cherts, with the remaining being cherts of unidentified type. These results are significantly reduced from the original dataset (Cascalheira et al., 2013), possibly due to no bucket coordinates being used.
From the 43 retouched tools, 11 different typologies were identified (without accounting for fragments). 
The typologies which are most frequently present in the studied sample are bifacial blanks and bifacial fragments (n=10). Two types of points were identified: shouldered points (n=3, including fragments) and tanged and winged points (n=4). Other frequent retouched typologies are retouched blanks (n=6), notches (n=5), splintered pieces (n=5) and endscrappers (n=4). Local types show a higher variability retouched typologies (9 types), although this may be related to the higher number of tools. In comparison, Type 6 shows only 5 different typologies.
